Diversity dynamics in Nymphalidae butterflies: Effect of phylogenetic uncertainty on diversification rate shift estimates

### 2014-12-19 MS_revised.tex
typo in acknowledgments
references in

Hostplant shifts have been invoked to be responsible for a great part of
the biodiversity of herbivorous insects \cite{mitter1988}. The study of the
evolution of hostplant use has spawned several theories explaining the
evolutionary interactions between plants and insects \cite{nyman2012}. The
``escape-and-radiate'' hypothesis \cite{ehrlich1964}, the 
``oscillation hypothesis''
\cite{janz2011,nylin2014} or ``diffuse cospeciation'' \cite{nyman2012} and the
``musical chairs hypothesis'' \cite{hardy2014}.

The butterfly family Nymphalidae has been an important taxon for
developing some of the mentioned hypotheses. Nymphalidae contains around
6000 species \cite{van_nieukerken2011},
and several members are considered model organisms
in evolutionary biology \cite{joron2006,willmott2006,brakefield2009}.
The family most likely originated
around 94 MYA in the mid Cretaceous. Diversification of the group began
in the Late Cretaceous and most major radiations (current subfamilies)
appeared shortly after the Cretaceous-Paleogene (K-Pg) boundary
\cite{heikkila2012}. Several studies have used time-calibrated phylogenies and
diversification models to reconstruct the evolutionary history of the
group to identify patterns of accelerated or decelerated diversification
of some Nymphalidae clades
\cite{heikkila2012,elias2009,fordyce2010,wahlberg2009}.
For example, it has been
suggested that climate change in the Oligocene and the subsequent
diversification of grasses has led to diversification of the subfamily
Satyrinae \cite{pena2008} due to the abundance of grasses over extensive
geographic areas (``resource abundance-dependent diversity dynamics''
hypothesis). Fordyce (2010) \cite{fordyce2010} found increased net diversification
rates in some Nymphalidae lineages after a major hostplant shift, which
appears to be in agreement with the ``escape-and-radiate'' model of
diversification \cite{ehrlich1964}.

Although it has been suggested that part of the great diversity of
Nymphalidae butterflies is a result of hostplant-insect dynamics, it is
necessary to use modern techniques to investigate whether the
diversification patterns of Nymphalidae are in agreement with the
theoretical predictions. It is necessary to test whether the overall
diversification pattern of Nymphalidae is congruent with events of
sudden diversification bursts due to hostplant shift or climatic events
\cite{nylin2014,ferrer2013}.

In this study, we used a time-calibrated genus-level phylogenetic
hypothesis for Nymphalidae butterflies \cite{wahlberg2009} to
investigate patterns
of diversification. We applied the statistical method
MEDUSA \cite{alfaro2009,harmon2011},
to study the diversification pattern of Nymphalidae
butterflies. MEDUSA fits likelihood models of diversification into a
time-calibrated tree and tests whether allowing increases or decreases
in speciation and extinction rates within the tree produces better fit
of the models. MEDUSA is able to take into account unsampled extant
species diversity during model fitting and it is normally applied to the
maximum clade credibility phylogenetic tree. Particularly, we wanted to
study the effects of phylogenetic uncertainty and by using the extended
MEDUSA method called MultiMEDUSA \cite{alfaro2009}. We also tested whether
hostplant association dynamics can explain the diversification patterns
of component Nymphalidae lineages by testing whether character states of
hostplant use affected the diversification pattern of those lineages
employing the method BiSSE as implemented in the R package
\texttt{diversitree} \cite{fitzjohn2012}.

@@

For analyses, we used the phylogenetic trees from the study of Wahlberg
et al. (2009) \cite{wahlberg2009} that were generated using DNA sequence data from
10 gene regions for 398 of the 540 valid genera in Nymphalidae. We
employed the maximum clade credibility tree (MCC tree) (Fig. 1) as well
as a random sample of 1000 trees from their BEAST run after burnin
\cite{wahlberg2009}. Their original BEAST run was for 40 million generations. We
used a burnin of 25 million generations and took a random sample of 1000
trees using Burntrees v.0.1.9 \url{http://www.abc.se/~nylander/} (supp.
mat. 5) in order to correct for phylogenetic uncertainty when performing
the diversification analyses.

Species richness data for Nymphalidae genera were compiled from several
sources including the specialist-curated lists on
\url{http://tolweb.org}, Lamas (2004) \cite{lamas2004} and curated lists of
Global Butterfly Names project (\url{http://www.ucl.ac.uk/taxome/gbn/}).
We assigned the species numbers of genera not included in the phylogeny
to the closest related genus that was included in Wahlberg et al. (2009)
\cite{wahlberg2009} study according to available phylogenetic studies 
\cite{matos2013,brower2010,kodandaramaiah2010,kodandaramaiah2010a,ortiz2013,desilva2010,freitas2004,pena2006,penz1999,silva2008,pena2011,pena2010},
taxonomical classification and morphological resemblance when no
phylogenies were available (supp. mat. 3).

Hostplant data for Nymphalidae species were compiled from several
sources including \cite{ackery1988}, HOSTS database (\url{http://bit.ly/YI7nwW}),
\cite{dyer2002,beccaloni2008,janzen2009} and others (supp. mat. 15--16)
for a total of 6586
hostplant records, including 428 Nymphalidae genera and 143 plant
families and 1070 plant genera. It was not possible to find any
hostplant data for 35 butterfly genera.

@@

Patterns of diversification in Nymphalidae were analyzed by using MEDUSA
version 093.4.33 \cite{alfaro2009} on the maximum clade credibility tree from
Wahlberg et al. (2009) \cite{wahlberg2009}. MEDUSA fits alternative birth-death
likelihood models to a phylogenetic tree in order to estimate changes in
net diversification rates along branches. MEDUSA estimates likelihood
and AIC scores for the simplest birth-death model, with two parameters,
the rates \texttt{r}: net diversification \texttt{=} speciation
(\texttt{b}) \texttt{-} extinction (\texttt{d}) and ($\varepsilon$:
relative extinction \texttt{= d/b}. The AIC scores of the two-parameter
model are then compared with incrementally more complex models until the
addition of parameters do not improve the AIC scores beyond a cutoff
value. MEDUSA finds the likelihood of the models after taking into
account branch lengths and number of species per lineage 
\cite{alfaro2009}. Most
studies using MEDUSA only run the method on a single tree, usually the
maximum clade credibility tree, which makes the assumption that this
tree is correct. We wanted to study the effects of phylogenetic
uncertainty on estimation of net diversification rate shifts and
therefore selected 1000 random genus-level trees from the posterior
distribution of the Bayesian run from Wahlberg et al., 2009 \cite{wahlberg2009}
(MultiMEDUSA, supp. mat. 5). We calculated a new MCC tree from the
selection of trees, ran MEDUSA and MultiMEDUSA on the selection of 1000
trees and summarized the estimated changes in net diversification rates
for nodes across all trees. Patterns of change in net diversification
rates are considered significant if they are found at the same node in
at least 90\% of the trees. We also expected to find similar values of
net diversification (\texttt{r = b - d}) and relative extinction rate
($\varepsilon$ \texttt{= d/b}) values across the 1000 trees for the
nodes where changes in diversification tempo occurs. We used the AICc
threshold of 7.8 units, as estimated by MEDUSA, as the limit for a
significantly better fit to select among increasingly complex
alternative models.

@@

As the MEDUSA and MultiMEDUSA approaches estimated an increase in net
diversification in the clade Ithomiini, we tested whether this pattern
can be explained by an increase in the birth-rate due to hostplant use
and performed analyses using the ``binary state speciation and
extinction'' \cite{maddison2007} Bayesian approach as implemented in the R package
\texttt{diversitree} \cite{fitzjohn2012}. The Markov Chain Monte Carlo algorithm
was run for 10000 generations discarding the first 7500 as burnin. The
Multiple State Speciation and Extinction approach (MuSSE) 
\cite{fitzjohn2012} is
designed to examine the joint effects of two or more traits on
speciation. Because most of Nymphalidae butterflies are restricted to
use one plant family as hostplant (232 genera use only one plant family
versus 176 that use more than one; based on our data in supp. mat. 15),
the character states can be coded as presence/absence, for which the
BiSSE analysis is better suited. BiSSE was designed to test whether a
binary character state has had any effect on increased net
diversification rate for a clade \cite{maddison2007}. We used our compiled data of
hostplant use to produce a binary dataset for the character ``feeding on
the plant family Solanaceae'' with two states (absence = 0; presence =
1) (supp. mat. 17). The family Solanaceae include the hostplants of the
diverse Ithomiini butterflies and closest relatives \cite{willmott2006}. Other
hostplant shifts were not tested for effect on net diversification rates
due to the low support for diversification shifts found in the MEDUSA
analyses. We analyzed the data using BiSSE employing the Markov Chain
Monte Carlo algorithm on the maximum clade credibility tree, taking into
account missing taxa by using the parameter ``sampling factor''
(\texttt{sampling.f}) in \texttt{diversitree}. We also used constrained
analyses forcing no effect of hostplant use on diversification and used
likelihood ratio tests to find out whether the hypothesis of effect on
diversification has a significantly better likelihood than the null
hypothesis (no effect). The analyses were run across a sample of 250
trees from the selected 1000 random trees from the posterior
distribution. The records of \emph{Vanessa} and \emph{Hypanartia}
feeding on Solanaceae \cite{beccaloni2008,scott1986} might be incorrect as it
is unlikely
that these species can be feeding on this plant family. We tested
whether coding these two genera as not feeding on Solanaceae affected
our results. We expected that coding these two genera as either absence
or presence would not have a significant effect on our results. It has
been shown that BiSSE performs poorly under certain conditions 
\cite{davis2013}.
However, our data has adequate number of taxa under analysis (more than
300 tips), adequate speciation bias (between 1.5x and 2.0x), character
state bias (around 8x) and extinction bias (around 4x) for the analysis
of Solanaceae hostplants. Thus, BiSSE is expected to produce robust
results \cite{davis2013}.

@@

We contrasted these results with those obtained from analyzing the data in
the software BAMM \cite{rabosky2014}. We found 
that the best model (best configuration) estimated 2 diversification shifts in
our MCC tree (posterior probability = 0.79) for the Ithomiini
and Satyrini (Fig. S1). Bayes factor (BF) values strongly 
favored this model in comparison with a constant rate model (no diversification
shifts, BF $>$ 1,884).

@@

The MEDUSA method has been used to infer changes in net diversification
rates in a phylogenetic tree. Since its publication \cite{alfaro2009} the results
of using MEDUSA on a single tree, the maximum clade credibility tree,
have been used for generation of hypotheses and discussion
\cite{heikkila2012,litman2011,ryberg2012}. However, different diversification
shifts and different
rates of diversification are found for certain lineages when
phylogenetic uncertainty was taken into account by using MEDUSA on a
random sample of trees from the posterior distribution of a Bayesian
run. We found that some diversification splits, estimated on the
Nymphalidae maximum clade credibility tree, were found with very low
probability in the random sample of 1000 trees from the posterior
distribution (Fig. 3, Table 2). We also found that, even though the
analyses estimated the same diversification splits on two or more trees,
the estimated net diversification rates could vary widely (Fig. 2). For
example, in our Nymphalidae trees, we found that the split for
\emph{Magneuptychia} and \emph{Caeruleuptychia} had a variation from
\texttt{r = 0.5234}, higher than the background net diversification
rate, to almost zero. This means that observed patterns and conclusions
can be completely contradictory depending on tree choice.

@@

Keith Brown suggested that feeding on Solanaceae was an important event
in the diversification of Ithomiini butterflies \cite{brown1987}. Ithomiini
butterflies are exclusively Neotropical and most species feed on
Solanaceae hostplants during larval stage \cite{willmott2006}.
Optimizations of the
evolution of hostplant use on phylogenies evidence a probable shift from
Apocynaceae to Solanaceae in the ancestor of the tribe 
\cite{willmott2006,brower2006}.
Fordyce (2010) \cite{fordyce2010} found that the Gamma statistics, a LTT plot of
an Ithomiini phylogeny and the fit of the density-dependent model of
diversification are consistent with a burst of diversification in
Ithomiini following the shift from Apocynaceae to Solanaceae.

@@

Solanaceae plants contain chemical compounds and it has been suggested
that the high diversity of Ithomiini is consistent with the
``escape-and-radiate scenario'' due to a shift onto Solanaceae 
\cite{fordyce2010}
and radiation scenarios among chemically different lineages of
Solanaceae plants \cite{willmott2006,brown1987}.
According to this theory, the shift from
Apocynaceae to Solanaceae allowed Ithomiini to invade newly available
resources due to a possible key innovation that allowed them cope with
secondary metabolites of the new hosts. Additional studies are needed to
identify the actual enzymes that Ithomiini species might be using for
detoxification of ingested food as they have been found in other
butterfly groups \cite{wheat2007}.

@@

Although the Solanaceae genera used by the Ithomiini clades are well
known \cite{willmott2006}, we do not have any understanding on the physiological
routes involved in the detoxification of Solanaceae compounds by the
several lineages of Ithomiini. We can speculate that older lineages
exploiting a novel toxic resource \cite{willmott2006,wahlberg2009}
might not be too efficient
in metabolizing plant toxins and that younger lineages are able to deal
with toxins more efficiently, so that host switching events within
Solanaceae are possible, which can lead to higher diversification.
Studies in \emph{Papilio} species have reported that detoxification
enzymes can become more efficient in metabolizing toxins than ancestral
configurations of the proteins, providing more opportunities for
hostplant switch \cite{li2003}. This might be the reason why the basal
Ithomiini subtribes Melinaeina and Mechanitina are so species-poor and
restricted to few Solanaceae hosts \cite{willmott2006}, while recent subtribes are
species-rich and have expanded their host range into several Solanaceae
lineages \cite{willmott2006}. It might be that the switch to feeding in Solanaceae
was an important event in the evolutionary history of Ithomiini, but the
actual radiation occurred after critical physiological changes (a
probable key innovation) allowed efficient detoxification of Solanaceae
toxins.

The diffuse cospeciation hypothesis predicts almost identical ages of
insects and their hostplants, while the ``resource abundance-dependent
diversity'' and the ``escape-and-radiate'' hypotheses state that insects
diversify after their hostplants \cite{nyman2012,ehrlich1964,janz2011}.
Wheat et al. (2007)
\cite{wheat2007} found strong evidence for a model of speciation congruent with
Ehrlich and Raven's hypothesis in Pieridae butterflies due to, in
addition to the identification of a key innovation, a burst of
diversification in glucosinolate-feeding taxa shortly afterwards (with a
lag of circa 10 MY). According to a recent dated phylogeny of
the Angiosperms \cite{bell2010}, the family Solanaceae split from its sister
group about 59 (49--68) MYA and diversification started (crown group
age) around 37 (29--47) MYA. Wahlberg et al. (2009) \cite{wahlberg2009} give the
ages for origin and diversification for Ithomiini at 45 (39--53) and 37
(32--43) MYA, respectively. Thus, current evidence shows that Solanaceae
and Ithomiini might have diversified around the same time, during the
Late Eocene and Oligocene, and this would be congruent with the diffuse
cospeciation hypothesis.

@@

Our MultiMEDUSA approach to perform analyses on the posterior
distribution of trees found strong support for an increase in net
diversification rate in the tribe Ithomiini and the genus
\emph{Charaxes}, and for a decrease in net diversification rate in the
subtribe Danaina. Due to phylogenetic uncertainty, we did not obtain
strong support for other diversification splits in Nymphalidae. Our
BiSSE analysis corroborated other studies in that the trait 
``feeding on Solanaceae'', or a codistributed character, was important in the
diversification of Ithomiini butterflies. However, by applying MEDUSA we
found that a critical character in the radiation of the group might have
appeared after the shift from Apocynaceae to Solanaceae. We also found
that the trait ``feeding on Apocynaceae'' is not responsible for the
slowdown of diversification in Danaina. Ithomiini and Solanaceae
diversified near simultaneously, which is in agreement with the diffuse
cospeciation hypothesis \cite{nyman2012,janz2011}.

@@

\section*{Figure legends}
\begin{description}
 \item {\bf Figure 1. Results of the MEDUSA analysis run on the maximum
clade credibility tree from Wahlberg et al. (2009)} Rate shifts were estimated for the
following nodes (besides the background rate): 2) Limenitidinae +
Heliconiinae, 3) \emph{Ypthima}, 4) \emph{Charaxes}, 5) Ithomiini in
part, 6) Satyrinae, 7) Coenonymphina 8) Phyciodina in part, 9) Danaini in
part, 10) \emph{Coenonympha}, 11) \emph{Caeruleuptychia} +
\emph{Magneuptychia}, 12) \emph{Callicore} + \emph{Diaethria}, 13)
Satyrina, 14) Mycalesina, 15) \emph{Pedaliodes}, 16) \emph{Dryas} +
\emph{Dryadula}, 17) \emph{Taenaris}, 18) Pseudergolinae. Circles on
nodes indicate the diversification shift number as found by MEDUSA.
Numbers next to circles indicate the posterior probability values for
such nodes.

\item {\bf Figure 2. Diversification rates for taxa estimated by MEDUSA on
the samples of 1000 random trees}

\item {\bf Figure 3. Results of MultiMEDUSA analysis showing the
probability of specific nodes being characterized by significant shifts
in diversification rate}

\item {\bf Figure 4. BiSSE analysis of diversification of nymphalids due
to feeding on Solanaceae hostplants} Speciation and net diversification
rates are significantly higher in Solanaceae feeders (speciation rate =
$\lambda1$, net diversification rate = $r1$.
\end{description}

\section*{Supporting Information Legends}
\begin{description}
 \item {\bf Supporting Information S01.nex.} MCC Nymphalidae tree from
\cite{wahlberg2009}

 \item {\bf Supporting Information S02.R} R script to run MEDUSA on the
MCC Nymphalidae tree from \cite{wahlberg2009}. This script removes the outgroup
taxa and loads the richness data for the tree terminals.

 \item {\bf Supporting Information S03.csv} Species richness for lineages
in Nymphalidae.

 \item {\bf Supporting Information S04.txt} Results of running MEDUSA on
the MCC tree.

 \item {\bf Supporting Information S05 1000 random trees no outgroups.zip} 1000 random trees from \cite{wahlberg2009}.

 \item {\bf Supporting Information S06.R} R script to run a MultiMEDUSA
analysis on 1000 random trees from \cite{wahlberg2009}.

 \item {\bf Supporting Information S07 multimedusa on 1000 trees.zip} Raw results from the MultiMEDUSA run on the random sample of trees from
the posterior distribution.

>>>>

Hostplant shifts have been invoked to be responsible for a great part of
the biodiversity of herbivorous insects \cite{mitter1988}. The study of the
evolution of hostplant use has spawned several theories explaining the
evolutionary interactions between plants and insects \cite{nyman2012}. The
``escape-and-radiate'' hypothesis \cite{ehrlich1964}, the ``oscillation
hypothesis'' \cite{janz2011, nylin2014} or ``diffuse cospeciation'' \cite{nyman2012} and the
``musical chairs hypothesis'' \cite{hardy2014}.

The butterfly family Nymphalidae has been an important taxon for
developing some of the mentioned hypotheses. Nymphalidae contains around
6000 species \cite{van_nieukerken2011}, and several members are considered model organisms
in evolutionary biology \cite{joron2006, willmott2006, brakefield2009}.
The family most likely originated
around 94 MYA in the mid Cretaceous. Diversification of the group began
in the Late Cretaceous and most major radiations (current subfamilies)
appeared shortly after the Cretaceous-Paleogene (K-Pg) boundary
\cite{heikkila2012}. Several studies have used time-calibrated phylogenies and
diversification models to reconstruct the evolutionary history of the
group to identify patterns of accelerated or decelerated diversification
of some Nymphalidae clades \cite{heikkila2012, elias2009, fordyce2010, wahlberg2009}.
For example, it has been
suggested that climate change in the Oligocene and the subsequent
diversification of grasses has led to diversification of the subfamily
Satyrinae \cite{pena2008} due to the abundance of grasses over extensive
geographic areas (``resource abundance-dependent diversity dynamics''
hypothesis): Fordyce (2010) \cite{fordyce2010} found increased net diversification
rates in some Nymphalidae lineages after a major hostplant shift, which
appears to be in agreement with the ``escape-and-radiate'' model of
diversification \cite{ehrlich1964}.

Although it has been suggested that part of the great diversity of
Nymphalidae butterflies is a result of hostplant-insect dynamics, it is
necessary to use modern techniques to investigate whether the
diversification patterns of Nymphalidae are in agreement with the
theoretical predictions. It is necessary to test whether the overall
diversification pattern of Nymphalidae is congruent with events of
sudden diversification bursts due to hostplant shift or climatic events
\cite{nylin2014, ferrer2013}.

In this study, we used a time-calibrated genus-level phylogenetic
hypothesis for Nymphalidae butterflies \cite{wahlberg2009} to investigate patterns
of diversification. We applied the statistical method MEDUSA
\cite{alfaro2009, harmon2011}, to study the diversification pattern of Nymphalidae
butterflies. MEDUSA fits likelihood models of diversification into a
time-calibrated tree and tests whether allowing increases or decreases
in speciation and extinction rates within the tree produces better fit
of the models. MEDUSA is able to take into account unsampled extant
species diversity during model fitting and it is normally applied to the
maximum clade credibility phylogenetic tree. Particularly, we wanted to
study the effects of phylogenetic uncertainty and by using the extended
MEDUSA method called MultiMEDUSA \cite{alfaro2009}. We also tested whether
hostplant association dynamics can explain the diversification patterns
of component Nymphalidae lineages by testing whether character states of
hostplant use affected the diversification pattern of those lineages
employing the binary speciation and extinction model (BiSSE) as
implemented in the R package \texttt{diversitree} \cite{fitzjohn2012}.

@@

For analyses, we used the phylogenetic trees from the study of Wahlberg
et al. (2009) \cite{wahlberg2009} that were generated using DNA sequence data from
10 gene regions for 398 of the 540 valid genera in Nymphalidae. We
employed the maximum clade credibility tree (MCC tree) (Fig. 1) as well
as a random sample of 1000 trees from their BEAST run after burnin
\cite{wahlberg2009}. Their original BEAST run was for 40 million generations. We
used a burnin of 25 million generations and took a random sample of 1000
trees using Burntrees v.0.1.9 \url{http://www.abc.se/~nylander/} (supp.
mat. 5) in order to account for phylogenetic uncertainty when performing
the diversification analyses.

Species richness data for Nymphalidae genera were compiled from several
sources including the specialist-curated lists on
\url{http://tolweb.org}, Lamas (2004) \cite{lamas2004} and curated lists of
Global Butterfly Names project (\url{http://www.ucl.ac.uk/taxome/gbn/}).
We assigned the species numbers of genera not included in the phylogeny
to the closest related genus that was included in Wahlberg et al. (2009)
\cite{wahlberg2009} study according to available phylogenetic studies 
\cite{matos2013, brower2010, kodandaramaiah2010, kodandaramaiah2010a,
ortiz2013, desilva2010, freitas2004, pena2006, penz1999, silva2008,
pena2011, pena2010},
taxonomical classification and morphological resemblance when no
phylogenies were available (supp. mat. 3).

Hostplant data for Nymphalidae species were compiled from several
sources including \cite{ackery1988}, HOSTS database (\url{http://bit.ly/YI7nwW}),
\cite{dyer2002, beccaloni2008, janzen2009} and others (supp. mat. 15--16) for a total of 6586
hostplant records, including 428 Nymphalidae genera and 143 plant
families and 1070 plant genera. It was not possible to find any
hostplant data for 35 butterfly genera.

@@

Patterns of diversification in Nymphalidae were analyzed by using MEDUSA
version 093.4.33 \cite{alfaro2009} on the maximum clade credibility tree from
Wahlberg et al. (2009) \cite{wahlberg2009}. MEDUSA fits alternative birth-death
likelihood models to a phylogenetic tree in order to estimate changes in
net diversification rates along branches. MEDUSA estimates likelihood
and AIC scores for the simplest birth-death model, with two parameters,
the rates \texttt{r}: net diversification \texttt{=} speciation
(\texttt{b}) \texttt{-} extinction (\texttt{d}) and (\(\varepsilon\):
relative extinction \texttt{= d/b}. The AIC scores of the two-parameter
model are then compared with incrementally more complex models until the
addition of parameters do not improve the AIC scores beyond a cutoff
value. MEDUSA finds the likelihood of the models after taking into
account branch lengths and number of species per lineage \cite{alfaro2009}. Most
studies using MEDUSA only run the method on a single tree, usually the
maximum clade credibility tree, which makes the assumption that this
tree is correct. We wanted to study the effects of phylogenetic
uncertainty on estimation of net diversification rate shifts and
therefore selected 1000 random genus-level trees from the posterior
distribution of the Bayesian run from Wahlberg et al., 2009 \cite{wahlberg2009}
(MultiMEDUSA, supp. mat. 5). We calculated a new MCC tree from the
selection of trees, ran MEDUSA and MultiMEDUSA on the selection of 1000
trees and summarized the estimated changes in net diversification rates
for nodes across all trees. Patterns of change in net diversification
rates are considered significant if they are found at the same node in
at least 90\% of the trees. We also expected to find similar values of
net diversification (\texttt{r = b - d}) and relative extinction rate
(\(\varepsilon\) \texttt{= d/b}) values across the 1000 trees for the
nodes where changes in diversification tempo occurs. We used the AICc
threshold of 7.8 units, as estimated by MEDUSA, as the limit for a
significantly better fit to select among increasingly complex
alternative models.

@@

As the MEDUSA and MultiMEDUSA approaches estimated an increase in net
diversification in the clade Ithomiini, we tested whether this pattern
can be explained by an increase in the birth-rate due to hostplant use
and performed analyses using the ``binary state speciation and
extinction'' \cite{maddison2007} Bayesian approach as implemented in the R package
\texttt{diversitree} \cite{fitzjohn2012}. The Markov Chain Monte Carlo algorithm
was run for 10000 generations discarding the first 7500 as burnin. The
Multiple State Speciation and Extinction approach (MuSSE) \cite{fitzjohn2012} is
designed to examine the joint effects of two or more traits on
speciation. Because most of Nymphalidae butterflies are restricted to
use one plant family as hostplant (232 genera use only one plant family
versus 176 that use more than one; based on our data in supp. mat. 15),
the character states can be coded as presence/absence, for which the
BiSSE analysis is better suited. BiSSE was designed to test whether a
binary character state has had any effect on increased net
diversification rate for a clade \cite{maddison2007}. We used our compiled data of
hostplant use to produce a binary dataset for the character ``feeding on
the plant family Solanaceae'' with two states (absence = 0; presence =
1) (supp. mat. 17). The family Solanaceae include the hostplants of the
diverse Ithomiini butterflies and closest relatives \cite{willmott2006}. Other
hostplant shifts were not tested for effect on net diversification rates
due to the low support for diversification shifts found in the MEDUSA
analyses. We analyzed the data using BiSSE employing the Markov Chain
Monte Carlo algorithm on the maximum clade credibility tree, taking into
account missing taxa by using the parameter ``sampling factor''
(\texttt{sampling.f}) in \texttt{diversitree}. We also used constrained
analyses forcing no effect of hostplant use on diversification and used
likelihood ratio tests to find out whether the hypothesis of effect on
diversification has a significantly better likelihood than the null
hypothesis (no effect). The analyses were run across a sample of 250
trees from the selected 1000 random trees from the posterior
distribution. The records of \emph{Vanessa} and \emph{Hypanartia}
feeding on Solanaceae \cite{beccaloni2008, scott1986} might be incorrect as it is unlikely
that these species can be feeding on this plant family. We tested
whether coding these two genera as not feeding on Solanaceae affected
our results. We expected that coding these two genera as either absence
or presence would not have a significant effect on our results. It has
been shown that BiSSE performs poorly under certain conditions \cite{davis2013}.
However, our data has adequate number of taxa under analysis (more than
300 tips), adequate speciation bias (between 1.5x and 2.0x), character
state bias (around 8x) and extinction bias (around 4x) for the analysis
of Solanaceae hostplants. Thus, BiSSE is expected to produce robust
results \cite{davis2013}.

@@

We contrasted these results with those obtained from analyzing the data
in the software BAMM \cite{rabosky2014}. We found that the best model (best
configuration) estimated 2 diversification shifts in our MCC tree
(posterior probability = 0.79) for the Ithomiini and Satyrini (Fig. S1).
Bayes factor (BF) values strongly favored this model in comparison with
a constant rate model (no diversification shifts, BF \(>\) 1,884).

@@

The MEDUSA method has been used to infer changes in net diversification
rates in a phylogenetic tree. Since its publication \cite{alfaro2009} the results
of using MEDUSA on a single tree, the maximum clade credibility tree,
have been used for generation of hypotheses and discussion
\cite{heikkila2012, litman2011, ryberg2012}. However, different diversification shifts and different
rates of diversification are found for certain lineages when
phylogenetic uncertainty was taken into account by using MEDUSA on a
random sample of trees from the posterior distribution of a Bayesian
run. We found that some diversification splits, estimated on the
Nymphalidae maximum clade credibility tree, were found with very low
probability in the random sample of 1000 trees from the posterior
distribution (Fig. 3, Table 2). We also found that, even though the
analyses estimated the same diversification splits on two or more trees,
the estimated net diversification rates could vary widely (Fig. 2). For
example, in our Nymphalidae trees, we found that the split for
\emph{Magneuptychia} and \emph{Caeruleuptychia} had a variation from
\texttt{r = 0.5234}, higher than the background net diversification
rate, to almost zero. This means that observed patterns and conclusions
can be completely contradictory depending on tree choice.

@@

Keith Brown suggested that feeding on Solanaceae was an important event
in the diversification of Ithomiini butterflies \cite{brown1987}. Ithomiini
butterflies are exclusively Neotropical and most species feed on
Solanaceae hostplants during larval stage \cite{willmott2006}. Optimizations of the
evolution of hostplant use on phylogenies evidence a probable shift from
Apocynaceae to Solanaceae in the ancestor of the tribe \cite{willmott2006, brower2006}.
Fordyce (2010) \cite{fordyce2010} found that the Gamma statistics, a LTT plot of
an Ithomiini phylogeny and the fit of the density-dependent model of
diversification are consistent with a burst of diversification in
Ithomiini following the shift from Apocynaceae to Solanaceae.

@@

Solanaceae plants contain chemical compounds and it has been suggested
that the high diversity of Ithomiini is consistent with the
``escape-and-radiate scenario'' due to a shift onto Solanaceae
\cite{fordyce2010}
and radiation scenarios among chemically different lineages of
Solanaceae plants \cite{willmott2006, brown1987}. According to this theory, the shift from
Apocynaceae to Solanaceae allowed Ithomiini to invade newly available
resources due to a possible key innovation that allowed them cope with
secondary metabolites of the new hosts. Additional studies are needed to
identify the actual enzymes that Ithomiini species might be using for
detoxification of ingested food as they have been found in other
butterfly groups \cite{wheat2007}.

@@

Although the Solanaceae genera used by the Ithomiini clades are well
known \cite{willmott2006}, we do not have any understanding on the physiological
routes involved in the detoxification of Solanaceae compounds by the
several lineages of Ithomiini. We can speculate that older lineages
exploiting a novel toxic resource \cite{willmott2006, wahlberg2009} might not be too efficient
in metabolizing plant toxins and that younger lineages are able to deal
with toxins more efficiently, so that host switching events within
Solanaceae are possible, which can lead to higher diversification.
Studies in \emph{Papilio} species have reported that detoxification
enzymes can become more efficient in metabolizing toxins than ancestral
configurations of the proteins, providing more opportunities for
hostplant switch \cite{li2003}. This might be the reason why the basal
Ithomiini subtribes Melinaeina and Mechanitina are so species-poor and
restricted to few Solanaceae hosts \cite{willmott2006}, while recent subtribes are
species-rich and have expanded their host range into several Solanaceae
lineages \cite{willmott2006}. It might be that the switch to feeding in Solanaceae
was an important event in the evolutionary history of Ithomiini, but the
actual radiation occurred after critical physiological changes (a
probable key innovation) allowed efficient detoxification of Solanaceae
toxins.

The diffuse cospeciation hypothesis predicts almost identical ages of
insects and their hostplants, while the ``resource abundance-dependent
diversity'' and the ``escape-and-radiate'' hypotheses state that insects
diversify after their hostplants \cite{nyman2012, ehrlich1964, janz2011}.
Wheat et al. (2007)
\cite{wheat2007} found strong evidence for a model of speciation congruent with
Ehrlich and Raven's hypothesis in Pieridae butterflies due to, in
addition to the identification of a key innovation, a burst of
diversification in glucosinolate-feeding taxa shortly afterwards (with a
lag of circa 10 MY). According to a recent dated phylogeny of the
Angiosperms \cite{bell2010}, the family Solanaceae split from its sister group
about 59 (49--68) MYA and diversification started (crown group age)
around 37 (29--47) MYA. Wahlberg et al. (2009) \cite{wahlberg2009} give the ages
for origin and diversification for Ithomiini at 45 (39--53) and 37
(32--43) MYA, respectively. Thus, current evidence shows that Solanaceae
and Ithomiini might have diversified around the same time, during the
Late Eocene and Oligocene, and this would be congruent with the diffuse
cospeciation hypothesis.

@@

Our MultiMEDUSA approach to perform analyses on the posterior
distribution of trees found strong support for an increase in net
diversification rate in the tribe Ithomiini and the genus
\emph{Charaxes}, and for a decrease in net diversification rate in the
subtribe Danaina. Due to phylogenetic uncertainty, we did not obtain
strong support for other diversification splits in Nymphalidae. Our
BiSSE analysis corroborated other studies in that the trait ``feeding on
Solanaceae'', or a codistributed character, was important in the
diversification of Ithomiini butterflies. However, by applying MEDUSA we
found that a critical character in the radiation of the group might have
appeared after the shift from Apocynaceae to Solanaceae. We also found
that the trait ``feeding on Apocynaceae'' is not responsible for the
slowdown of diversification in Danaina. Ithomiini and Solanaceae
diversified near simultaneously, which is in agreement with the diffuse
cospeciation hypothesis \cite{nyman2012, janz2011}.

@@

\section*{Figure legends}
{\bf Figure 1. Results of the MEDUSA analysis run on the maximum
clade credibility tree from Wahlberg et al. (2009)} Rate shifts were estimated for the
following nodes (besides the background rate): 2) Limenitidinae +
Heliconiinae, 3) \emph{Ypthima}, 4) \emph{Charaxes}, 5) Ithomiini in
part, 6) Satyrinae, 7) Coenonymphina 8) Phyciodina in part, 9) Danaini in
part, 10) \emph{Coenonympha}, 11) \emph{Caeruleuptychia} +
\emph{Magneuptychia}, 12) \emph{Callicore} + \emph{Diaethria}, 13)
Satyrina, 14) Mycalesina, 15) \emph{Pedaliodes}, 16) \emph{Dryas} +
\emph{Dryadula}, 17) \emph{Taenaris}, 18) Pseudergolinae. Circles on
nodes indicate the diversification shift number as found by MEDUSA.
Numbers next to circles indicate the posterior probability values for
such nodes.

{\bf Figure 2. Diversification rates for taxa estimated by MEDUSA on
the samples of 1000 random trees}

{\bf Figure 3. Results of MultiMEDUSA analysis showing the
probability of specific nodes being characterized by significant shifts
in diversification rate}

{\bf Figure 4. BiSSE analysis of diversification of nymphalids due
to feeding on Solanaceae hostplants} Speciation and net diversification
rates are significantly higher in Solanaceae feeders (speciation rate =
$\lambda1$, net diversification rate = $r1$.

\section*{Supporting Information Legends}
{\textbf{Supporting Information S01.nex.}} MCC Nymphalidae tree from
\cite{wahlberg2009}

{\textbf{Supporting Information S02.R}} R script to run MEDUSA on the
MCC Nymphalidae tree from \cite{wahlberg2009}. This script removes the outgroup
taxa and loads the richness data for the tree terminals.

{\textbf{Supporting Information S03.csv}} Species richness for lineages
in Nymphalidae.

{\textbf{Supporting Information S04.txt}} Results of running MEDUSA on
the MCC tree.

{\textbf{Supporting Information S05 1000 random trees no outgroups.zip}}
1000 random trees from \cite{wahlberg2009}.

{\textbf{Supporting Information S06.R}}~R script to run a MultiMEDUSA
analysis on 1000 random trees from \cite{wahlberg2009}.

{\textbf{Supporting Information S07 multimedusa on 1000 trees.zip}} Raw
results from the MultiMEDUSA run on the random sample of trees from the
posterior distribution.
